Formal Grammars as Types in Non-Commutative Linear-Non-Linear Type Theory

### 2024-07-10 POPL/paper.tex
Added comments and modified some paragraphs

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.
%
The final frontier for parsing research is the development of
\emph{formally verified} parser implementations, which due to the
pervasive nature of parsing would be a useful component of many
formally verified software systems: compilers and servers most
especially.
%

@@

The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{dep-lnl}.\max{had to stop here to take care of baby}

@@

\steven{The above paragraph maybe isn't the right thing to say, but I want
  something that ties the first paragraph to the contributions that follow}

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type thoery
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
include one to prove a metatheoretic canonicity theorem. Finally, in
\cref{sec:automata} we give a presentation of automata as grammars. With this
view in mind, we may encode parsing algorithms and execute them entirely
internal to our formalism. Our contributions are then:

@@

\steven{I initially wanted to say ``universal semantic domain to reason about
  parsing'', but these words have specific domain theory usage that aren't
  appropriate here, right?}

\steven{Check details for 2-cat of grammar formalisms}

@@

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users. Code in such a language can
become unnecessarily verbose when using Girard encodings to translate
high-level reasoning into low-level linear terms. To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

\subsection{Syntax}
\label{subsec:syntax}

Below we describe \theoryname~(\theoryabbv), an instance of an LNL theory with dependent types
to serve as a deductive
setting for formal grammar theory. \theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ to
specify and construct parsers.

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. These are all just as they appear in
\cite{krishnaswami_integrating_2015}. The treatment of the non-linear types is
pretty standard, so below we focus on the linear syntax.

@@

\subsection{Formal Grammars as Linear Types}
The linear typing judgment in our syntax takes on the following schematic form
$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
our syntax. The intended semantics of these linear types are formal grammars.
That is, the linear typing system is designed to syntactically reflect the
behavior of formal grammars. For this reason, we may often interchangeably use
the terms ``linear type'' and ``grammar''.

@@

$\Gamma$ represents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they do are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

@@

Most of the non-linear semantics is standard, so let us investigate in more
depth the linear semantics. The interpretation of each
connective is exactly what is suggested by the syntax, ude in
large part to the inspiration taken from the grammar interpretation
when designing \theoryabbv.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

@@

As a grammar,

\[
  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
\]

That is, $I$ maps strings to the proposition that they are the empty string. $I$
is clearly only inhabited by the empty string $\varepsilon$, for which the
outputted set contains a single parse tree.

@@

The grammar interpretation for characters is quite similar to that of $I$.
\[
  \sem {c} \gamma w = \{ () ~|~ w = c \}
\]

The grammar for the character $c$ likewise maps strings $w$ to the proposition that
they are equal to $c$.

@@

As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
linear types $A$, the following equalities hold:

\[ I \otimes A \equiv A \otimes I \equiv A \]

In the grammar semantics,

\[
  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
\]

The string $w$ matches $A \otimes B$
precisely when it may be split into two
pieces such that the left one matches $A$
and the right one matches $B$.

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.

\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]

The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.

\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]

That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.

@@

The grammar semantics of the linear product type is
indeed a dependent function out of the
semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
pair in $\Set$.

Note that even though we do not take the binary additive conjunction and
disjunction as primitive, we may define them via these LNL dependent types. In
particular, via a dependence on $\Bool$.

\[
  A \amp B := \LinPiTy b \Bool {C(b)}
\]
\[
  A \oplus B := \LinSigTy b \Bool {C(b)},
\]
where $C(\true) = A, C(\false) = B$.

\paragraph{Universal Type}
The universal type $\top$ may be formed in any context.

\[ \sem{\top} \gamma w = \{ \ast \}\]

Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants. The elimination for the empty type
witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.

\[ \sem {\bot} \gamma w = \emptyset \]

\paragraph{The $G$ Modality}
The $G$ modality provides the embedding of linear types in the empty context
into non-linear types. The introduction and elimination forms for $G A$ obey the
same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.

The left adjoint $F$ from non-linear types back to linear types may be defined
as,

\[
  F X := \LinSigTy a A I
\]

\steven{Elaborate on $G$ more}

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.

\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]

\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.

\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]

@@

@@

Under the Curry-Howard-Lambek correspondence, the programs of the simply-typed
$\lambda$-calculus may be viewed under any one of the three equivalent views: as
terms in a simple functional language, as proofs in an intuitionisitc
propositional logic, or as morphisms in a Cartesian closed category. In the
same vein, with the syntax provided, we aim to give a
categorical accounting of the semantics of \theoryabbv. Let us begin by paring back
the connectives in the language and only considering the subtheory induced by
regular expressions.

\steven{Added the above transition paragraph to handle putting the categorical semantics after
  the syntax. I think it sounds weird though}

By giving algebraic laws of manipulation and equivalence to regular expressions,
Kleene algebras are an important tool in the theory of regular
languages. More broadly, through various extensions, they serve as a
theoretical substrate to studying different kinds of formal languages.

@@

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we want to categorify the concept of
Kleene algebra and build on top of it in order to define an abstract
theory of parsing.  We start by defining \emph{Kleene categories}.

@@

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
\end{example}

This example illustrates how restricted Kleene algebras are in terms
of reasoning about languages. By only having at most one morphism
between objects, there is not a lot of information they can convey. In
this case, the only information you get is language containment. We
will see throughout this paper how the extra degrees of freedom
granted by having more morphisms give you more algebraic structure for
reasoning about languages.

@@

The proposed axioms are a direct translation of the Kleene algebra
axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family of initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: 1 + (\mu X.\, F_{A, 1}(X)) \otimes A \to
  (\mu X.\, F_{A, 1}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, 1}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, 1}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    1}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,1} \to B \lto Y$. We can show existence and
  uniqueness by showing that the diagram on top commutes if, and
  only if, the diagram on the bottom commutes:

% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
\[\begin{tikzcd}
	{B + B \otimes (\mu X.\, 1 + X \otimes A)} && {B + Y \otimes A} \\
	\\
	{B \otimes (\mu X.\, 1 + X \otimes A)} && Y \\
	{1 + (\mu X.\, 1 + X \otimes A)} && {1 + ( B \lto Y) \otimes A} \\
	\\
	{\mu X.\, 1 + X \otimes A} && {B \lto Y}
	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
	\arrow["{[f,g]}", from=1-3, to=3-3]
	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
	\arrow["h"', from=6-1, to=6-3]
	\arrow["{[f', g']}", from=4-3, to=6-3]
	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
\end{tikzcd}\]

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds.
\end{proof}

@@

For the rest of the paper we will use the presheaf category
$\Set^{\cat{\Sigma^*}}$ as a concrete model that will serve as our
guide when expanding abstract formalism presented in this section so
that it can handle more classes of languages beyond regular ones.

@@

\section{Other Models}
\label{sec:othermodels}
\subsection{Language Semantics}
Every grammar induces a language semantics. Also languages can be taken as a
propositionally truncated view of the syntax. Logical equivalence should induce
weak equivalence, and thus even give a syntactic way to reason about language equivalence.
\steven{TODO language semantics}

\subsection{Semantic Actions}
Returning to the problem of parsing, the output of a parse usually isn't the
literal parse tree. Rather, the output is the result of some \emph{semantic
  actions} ran on the parse tree, which usually serve to remove some syntactic
details that are unnecessary for later processing.

@@

>>>>

Parsing data from a serialized format is one of the most common tasks
in all of computing. Accordingly, this is a quite well-studied problem
spanning theoretical computer science and linguistics with such
milestones as Chomsky's hierarchy of grammars
\cite{chomThreeModels1956}, practical algorithms for parsing of
regular and context-free grammars, and variants
\cite{KNUTH1965607,Earley1970}, as well as implementations of
practical tools for the generation of efficient parsers
\cite{Johnsonyacc}.
%
The final frontier for parsing research is the development of
\emph{formally verified} parser implementations, which due to the
pervasive nature of parsing would be a useful component of many
formally verified software systems: compilers and servers most
especially.
%

@@

The type theory itself is a form of dependent linear-non-linear type
theory: the linear types can be interpreted as grammars, whereas the
non-linear types can be interpreted as ordinary sets. As in prior work
on dependent linear-non-linear types, the linear types are allowed to
depend on the non-linear types but not vice-versa \cite{dep-lnl}.

The substructural nature of \theoryabbv is well-aligned with the
requirements intrinsic to parsing and to the theory of formal
languages, where strings constitute a very clear notion of resource
that cannot be duplicated, reordered or dropped.

@@

\steven{The above paragraph maybe isn't the right thing to say, but I want
  something that ties the first paragraph to the contributions that follow}
\pedro{As a rule of thumb, I think that it's good to avoid subjective and
  pointed adjectives such as ``intuitive'', espcially if we don't have evidence
  that corroborates our claim. This is the perfect excuse for
an annoying reviewer to complain about our work.}

\paragraph{Contributions} This paper begins in \cref{sec:synindnotion} by
giving a universal semantic characterization of formal grammar.
From there, in \cref{sec:tt} we provide the syntax of \theoryname, a type thoery
that axiomatizes the underlying structures needed to conduct formal grammar
theory. Then in \cref{sec:categorify}, we characterize the abstract categorical
models of our syntax. Afterwards, we demonstrate several models of \theoryabbv,
including one to prove a metatheoretic canonicity theorem. Finally, in
\cref{sec:automata} we give a presentation of automata as grammars. With this
view in mind, we may encode parsing algorithms and execute them entirely
internal to our formalism. Our contributions are then:

@@

\steven{I initially wanted to say ``universal semantic domain to reason about
  parsing'', but these words have specific domain theory usage that aren't
  appropriate here, right?}
\pedro{If you want to keep ``semantic domain'', we could go for ``all-encompassing semantic domain''}
\steven{Check details for 2-cat of grammar formalisms}

@@

Unfortunately, programming in a linear term language is often
cumbersome and unintuitive for users. Code in such a language can
become unnecessarily verbose when using Girard encodings to translate
high-level reasoning into low-level linear terms.\pedro{What encodings are these?
We should add a citation here}
To alleviate this
concern, in 1995 Benton et al.\ proposed an alternative categorical
semantics of linear logic based on adjoint interactions between linear
and non-linear logics \cite{bentonMixedLinearNonlinear1995} ---
appropriately referred to as a \emph{linear-non-linear} (LNL)
system. This work is simply typed, so the boundary between linear and
non-linear subtheories is entirely characterized via a monoidal
adjunction $F \dashv G$ between linear terms and non-linear terms.

@@

In this section we go through a tour of \theoryabbv. We start by presenting its syntax.
In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
where the linear types have natural interpretation in terms of formal languages --- for instance,
the tensor corresponds to grammar concatenation. We conclude the section by showing how
classes of grammars correspond to certain subtype-theories inside \theoryabbv.

\pedro{Hopefully this serves as a segue between this and the syntax subsection}
\subsection{Syntax}
\label{subsec:syntax}

Below we describe \theoryname~(\theoryabbv), an instance of an LNL theory with dependent types
to serve as a deductive
setting for formal grammar theory. \theoryabbv axiomatizes the necessary structure underlying $\Set^{\String}$ \pedro{We should probably stick to either $\Set^{\String}$ or $\mathbf{Gr}$ from now on}to
specify and construct parsers.

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which also have a
natural grammar counterpart, which usually allows kinds of recursive
behavior such as the Kleene star. The treatment of the non-linear
types is pretty standard, so below we focus on the linear syntax.

@@

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}

The linear typing judgment in our syntax takes on the following schematic form
$\Gamma ; \Delta \vdash a : A$. First, $A$ represents a \emph{linear type} in
our syntax. The intended semantics of these linear types are formal grammars.
That is, the linear typing system is designed to syntactically reflect the
behavior of formal grammars. For this reason, we may often interchangeably use
the terms ``linear type'' and ``grammar''.

@@

$\Gamma$ represents a non-linear context, while $\Delta$ represents a
\emph{linear context} dependent on $\Gamma$. These linear contexts behave
substructurally. As stated earlier, they are linear --- they do not obey
weakening or contraction --- because a character is exhausted once read by a
parsing procedure. Moreover, the characters in strings appear in the order in
which we read them. We do not have the freedom to freely permute characters,
therefore any type theory that is used to reason about formal grammars ought to
omit the structural rule of exchange as well. This means that every variable
within $\Delta$ must be used \emph{exactly once} and \emph{in order of
  occurrence}. Thus, we can think of the linear contexts as an ordered list of
limited resources. Once a resource is consumed, we cannot make reference to it
again. Variables in a linear context then act like building
blocks for constructing patterns over strings.

@@

\pedro{The paragraphs above are a bit circular, what about the following alternative?}

Now that we have specified the syntactic aspect of \theoryabbv, we must justify
its connections to formal grammars in order to show that it is a sound methodology
for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
to our type theory.

Most of the non-linear semantics is standard \pedro{We should either cite something or add the non-linear semantics in the appendix}, so let us investigate in more
depth the linear semantics.

\begin{enumerate}
  \item A non-linear context $\Gamma$ denotes a set $\sem \Gamma$
  \item A non-linear type $\Gamma \vdash X : U_{i}$ denotes a family of sets
        $\sem X : \sem \Gamma \to \Set_{i}$
  \item A non-linear term $\Gamma \vdash e : X$ denotes a section
        $\sem e : \Pi(\gamma : \sem \Gamma)\sem{X} \gamma$ \pedro{we have not defined what sections are}
  \item Linear contexts $\Gamma \vdash \Delta$ and types
        $\Gamma \vdash A : L_{i}$ both denote families of grammars
        $\sem \Gamma \to \Gr_{i}$
  \item A linear term $\Gamma ; \Delta \vdash M : A$ denotes a family of parse
        transformers
        $\sem M : \Pi(\gamma : \sem \Gamma)\Pi(w : \String) \sem \Delta \gamma w \Rightarrow \sem M \gamma w$
\end{enumerate}

@@

As a grammar,
%
\[
  \sem {I} \gamma w = \{ () ~|~ w = \varepsilon \}
\]
%
That is, $I$ maps strings to the proposition that they are the empty string. $I$
is clearly only inhabited by the empty string $\varepsilon$, for which the
outputted set contains a single parse tree.

@@

The grammar interpretation for characters is quite similar to that of $I$.
%
\[
  \sem {c} \gamma w = \{ () ~|~ w = c \}
\]
%
The grammar for the character $c$ likewise maps strings $w$ to the proposition that
they are equal to $c$.

@@

As stated above, the type $I$ serves as the unit for $\otimes$. That is, for all
linear types $A$, the following equalities hold:
%
\[ I \otimes A \equiv A \otimes I \equiv A \]
%
In the grammar semantics,
%
\[
  \sem {A \otimes B} = \Sigma_{w_{1}, w_{2} : \String} (w_{1}w_{2} = w) \land \left(  \sem {A} \gamma w_{1} \times \sem {B} \gamma w_{2} \right)
\]
%
The string $w$ matches $A \otimes B$
precisely when it may be split into two
pieces such that the left one matches $A$
and the right one matches $B$.

\paragraph{Linear Function Types}
The monoidal structure provided by $\otimes$ is both left and right closed and
this is denoted by the left and right linear function types. Further,
for linear types $A$ and $B$ the left linear function type $A \lto B$
enjoys an elimination principle similar to function application or modus ponens.
%
\[
\inferrule{\Delta \vdash A \otimes A \lto B}{\Delta \vdash B}
\]
%
The interpretation of linear function types at a string $w$ is a linear function
that takes in parses of $A$ on a string $w_{a}$ and outputs parses of $B$ on the
string $w_{a} w$.
%
\[
  \sem{A \lto B} \gamma w = \Pi(w_a:\String) \left( A \gamma w_a \Rightarrow B\gamma (w_a w) \right)
\]
%
That is, strings match $A \lto B$ if when prepended with a parse of $A$ they
complete to parses of $B$. In this manner, the linear function types generalize
Brzozowksi's notion of derivative
\cite{brzozowskiDerivativesRegularExpressions1964}.
Brzozowski initially only gave an accounting of this operation for
generalized regular expressions, but later work from Might et al.\ demonstrates that the same
construction can be generalized to context free grammars
\cite{mightParsingDerivativesFunctional2011}. Here, via the linear function
types, the same notion of derivative extends to the grammars of \theoryabbv.
Note, to ensure that the linear function types do indeed generalize Brzozowski
derivative, we must include the equalities in \cref{fig:brzozowskideriv} as axioms.
\pedro{We should further explain why these equations are required, or at the very
least give some intuition}

@@

The grammar semantics of the linear product type is
indeed a dependent function out of the
semantics of $X$. Likewise, the grammar semantics of the linear dependent pair type is a dependent
pair in $\Set$.\pedro{The connection to grammar semantics needs to be better explained. For instance,
we should explain that disjunction corresponds to ``or'' of grammars}

Note that even though we do not take the binary additive conjunction and
disjunction as primitive, we may define them via these LNL dependent types. In
particular, via a dependence on $\Bool$.
%
\[
  A \amp B := \LinPiTy b \Bool {C(b)}
\]
\[
  A \oplus B := \LinSigTy b \Bool {C(b)},
\]
where $C(\true) = A, C(\false) = B$.

\paragraph{Universal Type}
The universal type $\top$ may be formed in any context.
%
\[ \sem{\top} \gamma w = \{ \ast \}\]
%
Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.

\paragraph{Empty Type}
The empty type $\bot$ has no inhabitants. The elimination for the empty type
witnesses the principle of explosion --- i.e.\ from a term of type $\bot$ we may
introduce a term $\mathsf{absurd}_{A} : A$ for any type $A$.
%
\[ \sem {\bot} \gamma w = \emptyset \]
%
\paragraph{The $G$ Modality}
The $G$ modality provides the embedding of linear types in the empty context
into non-linear types. The introduction and elimination forms for $G A$ obey the
same rules as given in \cite{bentonMixedLinearNonlinear1995} and \cite{krishnaswami_integrating_2015}.

The left adjoint $F$ from non-linear types back to linear types may be defined
as,
%
\[
  F X := \LinSigTy a A I
\]
%
\steven{Elaborate on $G$ more}

\pedro{Agreed :) maybe mention connections to persistent propositions from the
separation logic literature}

The semantics of $G A$ are exactly the semantics of $A$ at the empty word $\varepsilon$.
%
\[ \sem{G A} \gamma = \sem{A} \gamma \varepsilon \]
%
%
\paragraph{Recursive Types}
Recursive linear types may be defined via the least-fixed point operator $\mu$.
The grammar semantics of which are the fixed-point of sets induced by the
grammar semantics of $A$.
%
\[ \sem{\mu x. A} \gamma = \mu (x:\Gr_i). \sem{A}(\gamma,x) \]
%
\pedro{We should justify, even if briefly, the existence of this
  fixed-point}

@@

\pedro{This is an extremely important point! We should definitely make
  a reference to the Kleene algebra world, where these things do not
  hold. I think that it might even be a good idea to add the
  ``underlying trick'' to the main body of text. It's important to
  connect our work to existing ideas and concerns from the theory of
  formal languages.}

@@

An important part in the study of type theories is characterizing what
categorical structures they are modeling. In the case of \theoryabbv,
due to its intended connection to formal language theory, we use
Kleene algebras as a starting point in its categorical
semantics. Being posets, there is an immediate way of categorifying
them.

Kleene algebras are an important tool in the theory of regular languages
due to them being a bridge between algebraic reasoning and equivalence
of regular expressions. More broadly, through various extensions, they serve as a
theoretical substrate to studying different kinds of formal languages.

@@

The addition operation can be used to define the partial order
structure $a \leq b$ if, and only if, $a + b = b$. In the theory of
formal languages, this order structure can be used to model language
containment. In this section, we categorify the concept of Kleene
algebra and build on top of it in order to define an abstract theory
of parsing. We start by defining \emph{Kleene categories}.

@@

\begin{example}
  Every Kleene algebra, seen a posetal category, is a Kleene category.
  The product $\cdot$ is a monoidal product and the addition is a
  least-upper bound, which corresponds to a coproduct. Lastly, the
  axioms of the Kleene star have a direct correspondence to the
  coherence conditions postulated by the initial algebras of Kleene
  categories.
\end{example}

This example provides a neat categorical justification to how
restrictive Kleene algebras are in terms of reasoning about
languages. By only having at most one morphism between objects, there
is not a lot of information they can convey. In this case, the only
information you get is language containment. As demonstrated by
$\mathbf{Gr}$, the extra degrees of freedom granted by having more
morphisms gives you more algebraic structure for reasoning about
languages.

@@

As we have seen, the proposed axioms are a direct translation of the
Kleene algebra axioms to a categorical setting. Its most unusual aspect is the
axiomatization of the Kleene star as a family of initial algebras
satisfying certain isomorphisms. If the Kleene category $\cat{K}$ has
more structure, then these isomorphisms hold ``for free''.

\begin{theorem}
  \label{th:kleeneclosed}
  Let $\cat{K}$ be a Kleene category such that it is also monoidal
  closed.  Then, the initial algebras isomorphisms hold automatically.
\end{theorem}
\begin{proof}
  We prove this by the unicity (up-to isomorphism) of initial
  algebras. Let $[hd, tl]: I + (\mu X.\, F_{A, I}(X)) \otimes A \to
  (\mu X.\, F_{A, I}(X))$ be the initial algebra structure of $(\mu
  X.\, F_{A, I}(X))$ and consider the map $[hd, tl] : B + B \otimes
  (\mu X.\, F_{A, I}(X)) \otimes A \to B\otimes (\mu X.\, F_{A,
    I}(X))$.

  Now, let $[f,g] : B + A \otimes Y \to Y$ be an $F_{A,B}$-algebra and
  we want to show that there is a unique algebra morphism $h : \mu X.\, F_{A,I} \to B \lto Y$. We can show existence and
  uniqueness by showing that the diagram on top commutes if, and
  only if, the diagram on the bottom commutes:

% https://q.uiver.app/#q=WzAsOCxbMCwwLCJCICsgQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFsyLDAsIkIgKyBZIFxcb3RpbWVzIEEiXSxbMiwyLCJZIl0sWzAsMiwiQiBcXG90aW1lcyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDMsIjEgKyAoXFxtdSBYLlxcLCAxICsgWCBcXG90aW1lcyBBKSJdLFswLDUsIlxcbXUgWC5cXCwgMSArIFggXFxvdGltZXMgQSJdLFsyLDMsIjEgKyAoIEIgXFxsdG8gWSkgXFxvdGltZXMgQSJdLFsyLDUsIkIgXFxsdG8gWSJdLFswLDEsImlkICsgKGlkIFxcb3RpbWVzIGg7IGV2KSBcXG90aW1lcyBpZF9BIl0sWzEsMiwiW2YsZ10iXSxbMywyLCJpZCBcXG90aW1lcyBoOyBldiIsMl0sWzAsMywiW2lkIFxcb3RpbWVzIGgsIGlkIFxcb3RpbWVzIHRsXSIsMl0sWzQsNiwiaWQgKyAoaCBcXG90aW1lcyBYKVxcb3RpbWVzIGlkIl0sWzUsNywiaCIsMl0sWzYsNywiW2YnLCBnJ10iXSxbNCw1LCJbaGQsIHRsXSIsMl1d
\[\begin{tikzcd}
	{B + B \otimes (\mu X.\, I + X \otimes A)} && {B + Y \otimes A} \\
	\\
	{B \otimes (\mu X.\, I + X \otimes A)} && Y \\
	{I + (\mu X.\, I + X \otimes A)} && {I + ( B \lto Y) \otimes A} \\
	\\
	{\mu X.\, I + X \otimes A} && {B \lto Y}
	\arrow["{id + (id \otimes h; ev) \otimes id_A}", from=1-1, to=1-3]
	\arrow["{[f,g]}", from=1-3, to=3-3]
	\arrow["{id \otimes h; ev}"', from=3-1, to=3-3]
	\arrow["{[id \otimes h, id \otimes tl]}"', from=1-1, to=3-1]
	\arrow["{id + (h \otimes X)\otimes id}", from=4-1, to=4-3]
	\arrow["h"', from=6-1, to=6-3]
	\arrow["{[f', g']}", from=4-3, to=6-3]
	\arrow["{[hd, tl]}"', from=4-1, to=6-1]
\end{tikzcd}\]

  This equivalence follows by using the adjunction structure given
  by the monoidal closed structure of $\cat{K}$. A completely analogous
  argument for $G_{A,B}$ also holds. Furthermore, by generalizing the
  construction of \Cref{sec:formaltype}, we can also show that from
  the monoidal closed assumption it follows that $\mu X.\, F_{A, I}(X) \cong \mu X.\, G_{A, I}(X)$
\end{proof}

@@

@@

\pedro{I agree, then we should add the quick proof :) }

\section{Other Models}
\label{sec:othermodels}

One of the powers of type theories is that it can be profitable to interpret them
in various models. In this section, by using our just-defined Chomsky categories,
we show how other useful concepts from formal language theory can also be organized
as models of \theoryabbv. We illustrate this point by providing two examples that
are closely related to the theory of formal languages: language equivalence and
semantic actions. Furthermore, in order to justify how $\mathbf{Gr}$ relates to
more traditional notions of parsing, we define a glued model that proves a
canonicity property of grammar terms.

\subsection{Language Semantics}
Every grammar induces a language semantics. Also languages can be taken as a
propositionally truncated view of the syntax. Logical equivalence should induce
weak equivalence, and thus even give a syntactic way to reason about language equivalence.
\steven{TODO language semantics}

\subsection{Semantic Actions}
Returning to the problem of parsing, the output of a parse usually is not the
literal parse tree. Rather, the output is the result of some \emph{semantic
  actions} ran on the parse tree, which usually serve to remove some syntactic
details that are unnecessary for later processing.

@@

\pedro{This is a very nice opportunity of showing off the supremacy of denotational
  reasoning ;) We should probably prove the gluing lemma in the previous section
  and apply it here and in the canonicity section. The actual proof might have to be
  moved to the appendix, though}
